---
title: "STAT 341: Assignment 1"
subtitle: "Nichole Huang"
output:
  pdf_document:
    keep_tex: yes
    number_sections: no
    latex_engine: xelatex
  html_document:
    toc: yes
  word_document: default
urlcolor: blue
header-includes:
- \usepackage{multirow}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r, include = FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=70),tidy=TRUE)
```


## QUESTION 2: Basic R Calculations [10 points]

**(a)** 
```{r}
3^4

```
**(b)** 
```{r}
log(100, base=7)

```

**(c)** 
```{r}
sum(1 / (1:100)^2)

```
**(d)** 
```{r}
100 %% 7

```
**(e)**
\newline 
```{r}
A <- matrix(data = c(4,1,4,6,2,6,6,5,5,4,5,1),
            ncol = 4, byrow = FALSE)

max_row_sums <- max(apply(A, 1, sum))

mean_col_sums <- mean(apply(A, 2, sum))

cat("Maximum of the row sums:", max_row_sums, "\n")

cat("Mean of the column sums:", mean_col_sums, "\n")
``` 

$\;$

**(f)**
```{r}
# Define my own FUN function to count entries divisible by 5
count_divisible_by_5 <- function(row) {
  sum(row %% 5 == 0)  # Count how many elements are divisible by 5
}

divisible_entries <- apply(A, 1, count_divisible_by_5)

cat("Number of entries in each row divisible by 5:", divisible_entries, "\n")
```

$\;$

**(g)**
```{r}
nearest_neighbour <- function(v, x) {
  # calculate the absolute differences between each element in v and x
  differences <- sapply(v, function(element) abs(element - x))
  
  # Find the minimum difference
  min_diff <- min(differences)
  
  # Find the position of elements that are equal to the minimum difference
  nearest_pos <- which(differences == min_diff)
  
  return(nearest_pos)
}
    
    nearest_neighbour(v = c(4,1,4,6,2,6,6,5,5,4,5,1), x = 3)
```

\newpage

## QUESTION 3: Investigating the Weighted Average [13 points]

**(a)**

$$
\begin{aligned} 
a(\mathcal{P^*})&=a(y_1+b,...,y_N+b), \:b\in\mathbb{R}\\
&=\frac{\sum_{u\in\mathcal{P}}w_u(y_u+b)}{\sum_{u\in\mathcal{P}}w_u}\\
&=\frac{\sum_{u\in\mathcal{P}}w_uy_u}{\sum_{u\in\mathcal{P}}w_u}+\frac{\sum_{u\in\mathcal{P}}w_ub}{\sum_{u\in\mathcal{P}}w_u}\\
&=\frac{\sum_{u\in\mathcal{P}}w_uy_u}{\sum_{u\in\mathcal{P}}w_u}+\frac{b\sum_{u\in\mathcal{P}}w_u}{\sum_{u\in\mathcal{P}}w_u}\\
&=a(\mathcal{P})+b
\end{aligned}
$$
So, it's location equivariant.

$\;$

**(b)**

$$
\begin{aligned} 
a(\mathcal{P^*})
&=a(my_1,...,my_N),\:m>0\\
&=\frac{\sum_{u\in\mathcal{P}}w_u(my_u)}{\sum_{u\in\mathcal{P}}w_u}\\
&=\frac{m\sum_{u\in\mathcal{P}}w_uy_u}{\sum_{u\in\mathcal{P}}w_u}\\
&=m\times a(\mathcal{P})
\end{aligned}
$$
So, it's scale equivariant.

$\;$

**(c)**
$$
a(\mathcal{P^*})=a(\mathcal{P^k})=a(x_1,x_2,...,x_N),\:where\:x_i=ky_i,\:k\in\mathbb{N}\\
$$

$$
\begin{aligned} 
a(\mathcal{P^*})
&=\frac{\sum_{i\in\mathcal{P^*}}w_ix_i}{\sum_{i\in\mathcal{P^*}}w_i}\\
&=\frac{\sum_{u\in\mathcal{P}}w_u(ky_u)}{\sum_{u\in\mathcal{P}}kw_u}\\
&=\frac{k\sum_{u\in\mathcal{P}}w_uy_u}{k\sum_{u\in\mathcal{P}}w_u}\\
&=a(\mathcal{P})
\end{aligned}
$$
So, it's relication invariant.

$\;$

**(d)**
$$
\begin{aligned} 
SS(y,w;a(\mathcal{P}))
&=N[a(\mathcal{P^*})-a(\mathcal{P})]\\
&=N[\frac{{\sum_{u\in\mathcal{P}}w_uy_u}+wy}{{\sum_{u\in\mathcal{P}}w_u}+w} 
    -\frac{{\sum_{u\in\mathcal{P}}w_uy_u}}{{\sum_{u\in\mathcal{P}}w_u}}]\\
&=N[\frac{({\sum_{u\in\mathcal{P}}w_uy_u}+wy){\sum_{u\in\mathcal{P}}w_u}
        +({\sum_{u\in\mathcal{P}}w_u}+w){\sum_{u\in\mathcal{P}}w_uy_u}}
       {({\sum_{u\in\mathcal{P}}w_u}+w){\sum_{u\in\mathcal{P}}w_u}}]\\
&=\frac{Nw(y{\sum_{u\in\mathcal{P}}w_u}-{\sum_{u\in\mathcal{P}}w_uy_u})}
       {({\sum_{u\in\mathcal{P}}w_u}+w){\sum_{u\in\mathcal{P}}w_u}}
\end{aligned}
$$

$\;$

**(e)**
    
```{r}
set.seed(341)
N <- 1000
y_vals <- rexp(N)
w_vals <- runif(N)

# Calculate weighted mean of the original population
a_P <- weighted.mean(y_vals, w_vals)

# Sensitivity surface function SS(y, w; a(P))
sensitivity_surface <- function(y, w, y_vals, w_vals) {
  # Calculate weighted mean of the new population
  a_P_new <- weighted.mean(c(y_vals, y), c(w_vals, w))
  return (N * (a_P_new - a_P))
}

# Generate a grid of y and w values
y_grid <- seq(0, 10, length.out = 100)
w_grid <- seq(0, 3, length.out = 100)
yw_matrix <- outer(y_grid, w_grid, 
                   Vectorize(function(y, w) sensitivity_surface(y, w, y_vals, w_vals)))

# Plot the contour plot
contour(y_grid, w_grid, yw_matrix, xlab = "y", ylab = "w", 
        main = "Contour Plot of Sensitivity Surface SS(y, w; a(P))",
        col = hcl.colors(10, "Red-Blue"))
```

**(f)** Generally, as the injected y and w increase, the sensitivity of weighted average increases unboundedly. However, notice that when w = 0, sensitivity is 0; and when y is 1, no matter how w increases, the sensitivity remains 0.


\newpage 

## QUESTION 4: Weighted $k$-Nearest Neighbour Prediction [10 points]

$\;$

```{r}
dataq4 <- read.csv("q4data.csv")

# x: a vector of explanatory variate values for each of the units in P.
# y: a vector of response variate values for each of the units in P.
# k: the number of neighbors to consider when performing the weighted average prediction.
wknn <- function(x, y, k){
  N <- length(x)
  predictions <- numeric(N) # Initialize a vector of size N to store the predictions
  
  for (u in 1:N){
    
    # Calculate the distances between x_u and all the other x in the neighborhood
    distances <- abs(x[u]-x[-u])

    # Find the k nearest neighbors
    nearest_neighbors <- match(distances, sort(distances)) <= k

    # Calculate the weights
    weights <- exp(-distances[nearest_neighbors])
    
    # Calculate the weighted averages
    predictions[u] <- weighted.mean(y[-u][nearest_neighbors], weights)
  }
  return(predictions)
}

print(wknn(dataq4$x, dataq4$y, k=3))
```

\newpage


## QUESTION 5: Billboard 200™ [30 points] 

**(a)**
  
```{r, fig.width = 9}    
data_q5 <- read.csv("Billboard_200_Albums_2015-2024.csv")
par(mfrow = c(1, 3))

# Calculate the bin width using the Freedman-Diaconis rule
bin_width <- 2 * IQR(data_q5$Units) / (length(data_q5$Units)^(1/3))

# Create the sequence of breaks covering the full range of Units
breaks_seq <- seq(min(data_q5$Units), max(data_q5$Units) + bin_width, by = bin_width)

# Plot the histogram
hist(data_q5$Units, breaks = breaks_seq, 
     main = "Histogram of Units", 
     xlab = "Units Sold", 
     col = "lightblue", 
     border = "black")

# Plot the boxplot
boxplot(data_q5$Units, main = "Boxplot of Units", ylab = "Units Sold")

# Plot the quantile
qqnorm(data_q5$Units)
qqline(data_q5$Units, col = "red")
```

$\;$

**(b)**
```{r}
top_artists <- sort(table(data_q5$Artist), decreasing = TRUE)
head(top_artists, 3)
```

$\;$

**(c)**
```{r}
# Count the number of weeks each album was at number one
album_count <- table(data_q5$Album)

# Sort the albums by the number of weeks at number one, in descending order
sorted_album_count <- sort(album_count, decreasing = TRUE)

# Get the album with the most weeks at number one
top_album <- names(sorted_album_count)[1]   # Album name
top_album_artist <- data_q5$Artist[which(data_q5$Album == top_album)][1]  # Artist for the album
top_album_weeks <- sorted_album_count[1]    # Number of weeks at number one

cat(top_album, "by", top_album_artist,
    "with", top_album_weeks, "weeks at number one.")

```

$\;$

**(d)**
```{r}
# Find consecutive weeks for each album
album_runs <- rle(as.character(data_q5$Album))

# Find the most consecutive weeks
longest_run_index <- which.max(album_runs$lengths)
longest_run_album <- album_runs$values[longest_run_index]
longest_run_length <- album_runs$lengths[longest_run_index]

# Tie
longest_run_index_2 <- which.max(album_runs$lengths) + 1
longest_run_album_2 <- album_runs$values[longest_run_index_2]

# Find the artists for that albums
longest_run_artist <- data_q5$Artist[which(data_q5$Album == longest_run_album)][1]
longest_run_artist_2 <- data_q5$Artist[which(data_q5$Album == longest_run_album_2)][1]

cat(longest_run_album, 
    "by", longest_run_artist, 
    "with", longest_run_length, "consecutive weeks.")

cat(longest_run_album_2, 
    "by", longest_run_artist_2, 
    "with", longest_run_length, "consecutive weeks.")
```

$\;$

**(e)**
```{r}
# Calculate the total units sold by each album
total_units_per_album <- aggregate(Units ~ Album + Artist, data = data_q5, sum)

# Sort the albums by total units sold
sorted_total_units <- total_units_per_album[order(-total_units_per_album$Units), ]

# Find the album with the most units sold
top_units_album <- sorted_total_units[1, ]

cat(top_units_album$Album, 
    "by", top_units_album$Artist, 
    "with", top_units_album$Units, "total units sold.")
```

$\;$

**(f)**
```{r}
# Calculate the total units sold by each album for each year
total_units_per_album_year <- aggregate(Units ~ Year + Album + Artist, data = data_q5, sum)

# Find the album with the most units sold for each year
library(dplyr)
top_album_each_year <- total_units_per_album_year %>%
  group_by(Year) %>%
  top_n(1, Units)

top_album_each_year
```

$\;$

**(g)** 
```{r, fig.height = 6} 
x_g <- 1:507

# Create a scatterplot of `Units` over time
plot(x_g, y = data_q5$Units, xlim = c(0, 507),
     main = "Units Over Time", xlab = "Weeks", ylab = "Units Sold", 
     pch = 19, col = "lightblue", cex = 1)
points(x_g, y = data_q5$Units, pch = 1, cex = 1, col = "lightblue3")

# k-NN predictions with k = 2 and k = 20
k2_predictions <- wknn(x_g, data_q5$Units, k = 2)
k20_predictions <- wknn(x_g, data_q5$Units, k = 20)

# Smooth the lines
k2_smoothed <- smooth.spline(x_g, y = k2_predictions, spar=0.2)
k20_smoothed <- smooth.spline(x_g, y = k20_predictions, spar=0.2)

# Plot the lines
lines(k2_smoothed, col = "blue", lwd = 1.9)
lines(k20_smoothed, col = "violetred1", lwd = 1.9)

legend("topright", legend = c("2-NN", "20-NN"), col = c("blue", "violetred1"), lwd = 2)
```

$\;$

**(h)**
```{r}
# Calculate the absolute error for 2-NN and 20-NN predictions
true_values <- data_q5$Units
pred_2nn <- wknn(x_g, data_q5$Units, k = 2)
pred_20nn <- wknn(x_g, data_q5$Units, k = 20)

# Step 2: Calculate MAPE for 2-NN
mape_2nn <- mean(abs(true_values - pred_2nn))

# Step 3: Calculate MAPE for 20-NN
mape_20nn <- mean(abs(true_values - pred_20nn))

# Output the results
cat("MAPE for 2-NN is:", mape_2nn, "\n")
cat("MAPE for 20-NN is:", mape_20nn, "\n")

# Compare which k is better (smaller MAPE is better)
if (mape_2nn < mape_20nn) {
  cat("k = 2 provides better predictions.\n")
} else {
  cat("k = 20 provides better predictions.\n")
}

```

$\;$

**(i)**
```{r}   
# Define the transformation function
power_transform <- function(y, alpha) {
  if (sum(y <= 0) > 0) stop # y must be positive
  if (alpha == 0) {
    return(log(y))
  } else if (alpha > 0){
    return(y^alpha)
  } else {
    return(-y^alpha)
  }
}

# Apply transformations for alpha = -0.5, 0, 0.5
alpha_values <- c(-0.5, 0, 0.5)
mape_transformed <- c()


for (alpha in alpha_values) {
  
  # Apply the transformation
  y_transformed <- power_transform(data_q5$Units, alpha)
  
  # Apply weighted k-NN (using the best k from part (h))
  best_k <- if (mape_2nn < mape_20nn) 2 else 20
  pred_transformed <- wknn(x_g, y_transformed, k = best_k)
  
  # Inverse transformation
  if (alpha == 0) {
    pred_original <- exp(pred_transformed)
  } else {
    pred_original <- pred_transformed^(1 / alpha)
  }
  
  # Calculate MAPE
  mape_transformed[alpha + 1] <- mean(abs(data_q5$Units - pred_original))
}

# Output the best transformation based on MAPE
best_alpha <- alpha_values[which.min(mape_transformed)]
cat("Best transformation is for alpha =", best_alpha, "with MAPE =", min(mape_transformed), "\n")
```
Applying weighted k-NN to symmetric data appears to make a significant difference compared to skewed data. The reduction in MAPE indicates that the transformation helped stabilize the variance and improved the model's ability to predict. This suggests that using symmetric data allows the weighted k-NN algorithm to function more effectively, possibly because it can reduce the influence of outliers.
    
