---
title: 'STAT 341: Assignment 3'
subtitle: 'DUE: Tuesday December 3, 2024 by 5:00pm ET'
output:
  pdf_document:
    extra_dependencies:
    - bm
    - bbm
    - amsmath
    - physics
  html_document:
    df_print: paged
urlcolor: blue
---
```{r setup, include = FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  tidy = "styler"
)
```

$\;$
$\;$
$\;$
$\;$

## NOTES

Your assignment must be submitted by the due date listed at the top of this document, and it must be submitted electronically in .pdf format via Crowdmark. This means that your responses for different questions should begin on separate pages of your .pdf file. Note that your .pdf solution file must have been generated by R Markdown. Additionally:

* For mathematical questions: your solutions must be produced by LaTeX (from within R Markdown). Neither screenshots nor scanned/photographed handwritten solutions will be accepted -- these will receive zero points.

* For computational questions: R code should always be included in your solution (via code chunks in R Markdown). If code is required and you provide none, you will receive zero points.
   + **Exception:** any functions defined in the lecture notes can be loaded using `echo=FALSE` but any other code chunks should have `echo=TRUE`. e.g., the code chunk loading `mixRandomly()` can use `echo=FALSE` but chunks that call `mixRandomly()` should have `echo=TRUE`.
   
* For interpretation questions: plain text (within R Markdown) is required. Text responses embedded as comments within code chunks will not be accepted.

Organization and comprehensibility is part of a full solution. Consequently, points will be deducted for solutions that are not organized and incomprehensible. Furthermore, if you submit your assignment to Crowdmark, but you do so incorrectly in any way (e.g., you upload your Question 2 solution in the Question 1 box), you will receive a 5% deduction (i.e., 5% of the assignment’s point total will be deducted from your point total).

\newpage


## QUESTION 2: Variance of the Horvitz-Thompson Estimator [12 points]

**(a)**
$\sum_{u\in\mathcal{P}}\pi_u = n$: When sampling without replacement, each unit in the sample is unique, so the total number of distinct units is $n = \sum_{u\in\mathcal{P}}D_u$. Taking the expectation on both sides yields
$$
\begin{aligned} 
E[n] 
&= E[\sum_{u\in\mathcal{P}}D_u] \\
&n = \sum_{u\in\mathcal{P}}E[D_u]  \\
&n = \sum_{u\in\mathcal{P}}\pi_u
\end{aligned}
$$

$\sum_{v\in\mathcal{P}}\pi_{uv}=n\pi_u$: Since each unit $u$ is selected in $n$ different ways, and each way pairs $u$ with other $n-1$ units. So we have 
$$
\sum_{v\in\mathcal{P}}\pi_{uv} = (n-1) \times \pi_u + \pi_{uu} = n\pi_u - \pi_u + \pi_u = n\pi_u
$$
$\;$

**(b)**
$$
\begin{aligned}
Var[\tilde a_{HT}(S)]
&=-\frac{1}{2} \sum_{u\in\mathcal{P}}\sum_{v\in\mathcal{P}}
  (\pi_{uv} - \pi_{u}\pi_{v}) (\frac{t_u}{\pi_{u}} - \frac{t_v}{\pi_{v}})^2 \\
&=-\frac{1}{2} \sum_{u\in\mathcal{P}}\sum_{v\in\mathcal{P}}
  (\pi_{uv} - \pi_{u}\pi_{v}) 
  (\frac{t_u^2}{\pi_{u}^2} + \frac{t_v^2}{\pi_{v}^2} - 
   2\frac{t_u}{\pi_{u}}\frac{t_v}{\pi_{v}}) \\
&=-\frac{1}{2} \sum_{u\in\mathcal{P}}\sum_{v\in\mathcal{P}}
  (\pi_{uv} - \pi_{u}\pi_{v}) \frac{t_u^2}{\pi_{u}^2} 
  -\frac{1}{2} \sum_{u\in\mathcal{P}}\sum_{v\in\mathcal{P}}
  (\pi_{uv} - \pi_{u}\pi_{v}) \frac{t_v^2}{\pi_{v}^2}
  +\sum_{u\in\mathcal{P}}\sum_{v\in\mathcal{P}}
  (\pi_{uv} - \pi_{u}\pi_{v}) \frac{t_u}{\pi_{u}}\frac{t_v}{\pi_{v}} \\
&=\sum_{u\in\mathcal{P}}\sum_{v\in\mathcal{P}}
  (\pi_{uv} - \pi_{u}\pi_{v}) \frac{t_u}{\pi_{u}}\frac{t_v}{\pi_{v}}
\end{aligned}
$$

$\;$

**(c)**
```{r}
var_ht <- function(t_u, pi_u, pi_uv) {
  delta <- pi_uv - outer(pi_u, pi_u) 
  ht_terms <- outer(t_u / pi_u, t_u / pi_u)
  estimateVar <- sum(delta * ht_terms)
  return(estimateVar)
}

var_ht_syg <- function(t_u, pi_u, pi_uv) {
  delta <- pi_uv - outer(pi_u, pi_u) 
  syg_terms <- outer(t_u / pi_u, t_u / pi_u, "-")^2 
  estimateVar <- -0.5 * sum(delta * syg_terms)
  return(estimateVar)
}
```

**(d)**
```{r}
test_descr <- expand.grid(
  is_N_one = c(FALSE, TRUE),   # N > 1 or N = 1
  is_n_one = c(FALSE, TRUE),   # n > 1 or n = 1
  is_n_N   = c(FALSE, TRUE)    # n < N or n = N
)
test_descr
# eliminate impossible settings
rem_id <- c(
  2, # N = 1, n > 1, n < N
  4, # N = 1, n = 1, n < N
  6, # N = 1, n > 1, n = N
  7  # N > 1, n = 1, n = N
)
test_descr <- test_descr[-rem_id,]
test_descr # impossible settings removed
```

```{r}
library(testthat)

perm <- function(x) {
  if (length(x) == 1) {
    return(x)
  } else {
    res <- matrix(ncol = 0, nrow = length(x))
    for (i in seq_along(x)) {
      res <- cbind(res, rbind(x[i], Recall(x[-i])))
    }
    return(res)
  }
}

get_joint_inclusion <- function(size, prob) {
  # all unordered samples
  N <- length(prob)
  all_samples <- combn(N, size)
  n_samples <- ncol(all_samples)
  # all permutations to get ordered samples
  all_perms <- as.matrix(perm(1:size))
  # compute probability of each unordered sample
  all_prob <- rep(0, n_samples)
  for(i in 1:ncol(all_perms)) {
    cum_prob <- 1 # cumulative probability of ordered sample
    rem_prob <- rep(sum(prob), n_samples) # remaining probabilty
    for(j in 1:size) {
      pr <- prob[all_samples[all_perms[j,i],]]
      cum_prob <- cum_prob * (pr / rem_prob)
      rem_prob <- rem_prob - pr
    }
    all_prob <- all_prob + cum_prob
  }
  # compute the joint inclusion matrix
  pi_mat <- matrix(0, N, N)
  for(i in 1:N) {
    for(j in 1:N) {
      # which samples contain (i, j)
      id <- (i == all_samples) + (j == all_samples)
      id <- colSums(id) == 2
      pi_mat[i, j] <- sum(all_prob[id])
    }
  }
  pi_mat
}

# Find inclusion probabilities
get_joint_inclusion <- function(N, n) {
  pi_u <- runif(N, 0.1, 1)
  pi_uv <- outer(pi_u, pi_u, function(u, v) u * v)
  list(pi_u = pi_u, pi_uv = pi_uv)
}

# Unit test
test_that("var_ht matches var_ht_syg for any sampling scheme", {
  for (i in 1:nrow(test_descr)) {
    case <- test_descr[i, ]
    N <- if (case$is_N_one) 1 else sample(2:8, 1)
    n <- if (case$is_n_one) 1 else if (case$is_n_N) N else sample(1:(N - 1), 1)
    data <- get_joint_inclusion(N, n)
    t_u <- runif(N, 1, 10)
    expect_equal(
      var_ht(t_u, data$pi_u, data$pi_uv),
      var_ht_syg(t_u, data$pi_u, data$pi_uv),
      tolerance = 1e-8
    )
  }
})
```


\newpage

## DATA FOR QUESTIONS 3-5

The data in `NBA_2024_Shots.csv` includes $27$ variates about all $218,701$ shots taken in the 2023-24 National Basketball Association regular season. This data was scraped from [nba.com](https://www.nba.com/) and is available on [Kaggle](https://www.kaggle.com/datasets/mexwell/nba-shots/)^[If you're interested in this data, the Kaggle webpage contains equivalent datasets for all seasons between 2004-2024.]. An abridged data dictionary follows. For every shot, the following variates will be of interest.

| Variable               | Description                                                 |
|------------------------|-------------------------------------------------------------|
| `TEAM_NAME`            | The name of the team of the player who took the shot.       |
| `TEAM_NAME_ABBREV`     | A 3-letter abbreviation for the name of the team of the player who took the shot.       |
| `PLAYER_NAME`          | The name of the player who took the shot.                   |
| `HOME_TEAM`            | The home team in the game in which the shot was taken. This is recorded as the team's 3-letter abbreviation.      |
| `AWAY_TEAM`            | The away team in the game in which the shot was taken. This is recorded as the team's 3-letter abbreviation.       |
| `SHOT_MADE`            | A boolean variate indicating whether the shot was successful (`TRUE`) or unsuccessful (`FALSE`). |
| `LOC_X`                | The $x$ coordinate of the shot in the $(x,y)$ plane of the court. This dimension corresponds to the width of the court, which is recorded in feet and coded between `(-25,25)`.|
| `LOC_Y`                | The $y$ coordinate of the shot in the $(x,y)$ plane of the court. This dimension corresponds to the length of the court, which is recorded in feet and coded between `(0,94)`.|


Below is a visualization of the court, which should help to interpret `LOC_X` and `LOC_Y`. Note that this was drawn with the `draw_court()` function available in Appendix B. The target hoop is the one located at (0,4).

```{r, echo=FALSE, fig.height=5.5, fig.align='center'}
draw_court <- function(add = FALSE, lwd = 1, col = "black", axes = FALSE) {
  if(!add) {
    plot(0,0, col = "white", asp = 1,
         xlim = c(-30,30), xaxt = "n", xlab = "",
         ylim = c(0,95), yaxt = "n", ylab = "")
    if(axes) {
      axis(side = 1, cex.axis = 0.75)
      axis(side = 2, cex.axis = 0.75)
    }
  }
  # baselines
  segments(x0 = -25, y0 = 0, x1 = 25, y1 = 0, lwd = lwd, col = col)
  segments(x0 = -25, y0 = 94, x1 = 25, y1 = 94, lwd = lwd, col = col)
  # sidelines
  segments(x0 = -25, y0 = 0, x1 = -25, y1 = 94, lwd = lwd, col = col)
  segments(x0 = 25, y0 = 0, x1 = 25, y1 = 94, lwd = lwd, col = col)
  # half court
  segments(x0 = -25, y0 = 47, x1 = 25, y1 = 47, lwd = lwd, col = col)
  lines(x = seq(-6,6,0.01), y = sqrt(36-seq(-6,6,0.01)^2)+47, lwd = lwd, col = col)
  lines(x = seq(-6,6,0.01), y = -sqrt(36-seq(-6,6,0.01)^2)+47, lwd = lwd, col = col)
  # 3-point arcs
  lines(x = seq(-22,22,0.01), y = sqrt(23.9^2-seq(-22,22,0.01)^2)+4.7, lwd = lwd, col = col)
  segments(x0 = -22, y0 = 0, x1 = -22, y1 = 14, lwd = lwd, col = col)
  segments(x0 = 22, y0 = 0, x1 = 22, y1 = 14,lwd = lwd, col = col)
  lines(x = seq(-22,22,0.01), y = -sqrt(23.9^2-seq(-22,22,0.01)^2)+89.3, lwd = lwd, col = col)
  segments(x0 = -22, y0 = 80, x1 = -22, y1 = 94, lwd = lwd, col = col)
  segments(x0 = 22, y0 = 80, x1 = 22, y1 = 94, lwd = lwd, col = col)
  # keys
  lines(x = seq(-6,6,0.01), y = sqrt(36-seq(-6,6,0.01)^2)+15, lwd = lwd, col = col)
  lines(x = seq(-6,6,0.01), y = -sqrt(36-seq(-6,6,0.01)^2)+15, lwd = lwd, lty = 3, col = col)
  segments(x0 = -8, y0 = 0, x1 = -8, y1 = 15, lwd = lwd, col = col)
  segments(x0 = 8, y0 = 0, x1 = 8, y1 = 15, lwd = lwd, col = col)
  segments(x0 = -8, y0 = 15, x1 = 8, y1 = 15, lwd = lwd, col = col)
  lines(x = seq(-6,6,0.01), y = -sqrt(36-seq(-6,6,0.01)^2)+79, lwd = lwd, col = col)
  lines(x = seq(-6,6,0.01), y = sqrt(36-seq(-6,6,0.01)^2)+79, lwd = lwd, lty = 3, col = col)
  segments(x0 = -8, y0 = 94, x1 = -8, y1 = 79, lwd = lwd, col = col)
  segments(x0 = 8, y0 = 94, x1 = 8, y1 = 79, lwd = lwd, col = col)
  segments(x0 = -8, y0 = 79, x1 = 8, y1 = 79, lwd = lwd, col = col)
  # baskets
  points(x = 0, y = 4, pch = 16, col = col)
  segments(x0 = -2, y0 = 3, x1 = 2, y1 = 3, lwd = lwd, col = col)
  points(x = 0, y = 90, pch = 16, col = col)
  segments(x0 = -2, y0 = 91, x1 = 2, y1 = 91, lwd = lwd, col = col)
}
draw_court(axes = TRUE)
```

\newpage

## QUESTION 3: Sampling Distributions and Confidence Intervals [22 points]

**(a)**
```{r}
      nba <- read.csv("NBA_2024_Shots.csv", header = TRUE)
      set.seed(341)
      srswor_indices <- sample(x = nrow(nba), size = 1000)
      srswor_sample <- nba[srswor_indices,]
```
      i. 
```{r}
N <- nrow(nba)
n <- 1000
srswor_sample$shot_prob <- n / N

ht_estimate <- 100 / N * sum(srswor_sample$SHOT_MADE / srswor_sample$shot_prob)
ht_estimate
```

      ii. 
```{r}
# Horvitz-Thompson CI
pi_u <- rep(n / N, n)
pi_uv <- matrix((n * (n - 1))/(N * (N - 1)), nrow = n, ncol = n)
t_u <- 100 / N * srswor_sample$SHOT_MADE
diag(pi_uv) <- pi_u

est_var_ht <- function(t_u, pi_u, pi_uv) {
  delta <- pi_uv - outer(pi_u, pi_u)
  var_hat <- sum((delta/pi_uv) * outer(t_u/pi_u, t_u/pi_u))
  return(var_hat)
}

var_est <- est_var_ht(t_u, pi_u, pi_uv)
c(ht_estimate - 1.96 * sqrt(var_est), ht_estimate + 1.96 * sqrt(var_est))

# Bootstrap CI
set.seed(341)
B <- 1000
boot_estimates <- replicate(B, {
  boot_sample <- srswor_sample[sample(1:n, n, replace = TRUE), ]
  sum(boot_sample$SHOT_MADE / boot_sample$shot_prob) * 100 / N
})
var_boot <- var(boot_estimates)
c(ht_estimate - 1.96 * sqrt(var_boot), ht_estimate + 1.96 * sqrt(var_boot))

# Percentile CI
quantile(boot_estimates, probs = c(0.025, 0.975))
```

$\;$

**(b)**

    i. 
```{r}
library(ggplot2)

true_fg <- 100 * mean(nba$SHOT_MADE)  # True FG% from the population
sample_sizes <- c(50, 100, 500, 1000)
B <- 500  # Number of bootstrap resamples per sample
num_simulations <- 500  # Number of SRSWOR samples per n

results <- lapply(sample_sizes, function(n) {
  set.seed(341)
  bias <- numeric(num_simulations)
  mse <- numeric(num_simulations)
  var_ht <- numeric(num_simulations)
  coverage_normal <- numeric(num_simulations)
  coverage_bootstrap <- numeric(num_simulations)
  coverage_percentile <- numeric(num_simulations)
  
  for (i in 1:num_simulations) {
    # Draw random sample
    indices <- sample(x = 1:N, size = n, replace = FALSE)
    sample_n <- nba[indices, ]
    
    # HT estimate
    pi_u <- rep(n / N, n)  
    pi_uv <- matrix((n * (n - 1))/(N * (N - 1)), nrow = n, ncol = n)
    diag(pi_uv) <- pi_u
    t_u <- 100 / N * sample_n$SHOT_MADE
    ht_est <- sum(t_u / pi_u)
    
    # Variance using est_var_ht
    var_hat <- est_var_ht(t_u, pi_u, pi_uv)
    
    # Normal-theory CI
    ci_normal <- c(ht_est - 1.96 * sqrt(var_hat), ht_est + 1.96 * sqrt(var_hat))
    
    # Bootstrap CIs
    boot_estimates <- replicate(B, {
      boot_sample <- sample_n[sample(1:n, n, replace = TRUE), ]
      t_u_boot <- boot_sample$SHOT_MADE
      sum(t_u_boot / pi_u) * 100 / N
    })
    ci_boot <- c(ht_est - 1.96 * sqrt(var(boot_estimates)), 
                 ht_est + 1.96 * sqrt(var(boot_estimates)))
    ci_percentile <- quantile(boot_estimates, probs = c(0.025, 0.975))
    
    # Store metrics
    bias[i] <- ht_est - true_fg
    mse[i] <- (ht_est - true_fg)^2
    var_ht[i] <- var_hat
    coverage_normal[i] <- (ci_normal[1] <= true_fg) & (ci_normal[2] >= true_fg)
    coverage_bootstrap[i] <- (ci_boot[1] <= true_fg) & (ci_boot[2] >= true_fg)
    coverage_percentile[i] <- (ci_percentile[1] <= true_fg) & (ci_percentile[2] >= true_fg)
  }
  # Summary metrics
  list(
    n = n,
    bias = mean(bias),
    variance = mean(var_ht),
    mse = mean(mse),
    coverage_normal = mean(coverage_normal),
    coverage_bootstrap = mean(coverage_bootstrap),
    coverage_percentile = mean(coverage_percentile)
  )
})

# Combine results into a data frame
plot_data <- do.call(rbind, lapply(results, function(res) {
  data.frame(
    n = res$n,
    Bias = res$bias,
    Variance = res$variance,
    MSE = res$mse,
    Coverage_Normal = res$coverage_normal,
    Coverage_Bootstrap = res$coverage_bootstrap,
    Coverage_Percentile = res$coverage_percentile
  )
}))

# Plot Bias, Variance, MSE, and Coverage
bias_plot <- 
  ggplot(plot_data, aes(x = n)) +
  geom_line(aes(y = Bias), color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(y = "Bias", x = "Sample Size (n)", title = "Bias") +
  ylim(-5, 5) + 
  theme(plot.title = element_text(hjust = 0.5))

variance_plot <-
  ggplot(plot_data, aes(x = n)) +
  geom_line(aes(y = Variance), color = "red") +
  labs(y = "Variance", x = "Sample Size (n)", title = "Variance") +
  theme(plot.title = element_text(hjust = 0.5))

mse_plot <-
  ggplot(plot_data, aes(x = n)) +
  geom_line(aes(y = MSE), color = "purple") +
  labs(y = "MSE", x = "Sample Size (n)", title = "MSE") +
  theme(plot.title = element_text(hjust = 0.5))

coverage_normal_plot <-
  ggplot(plot_data, aes(x = n)) +
  geom_line(aes(y = Coverage_Normal), color = "darkblue") +
  geom_hline(yintercept = 0.95, linetype = "dashed") +
  labs(y = "Coverage", x = "Sample Size (n)", title = "Normal CI") +
  ylim(0, 1) +
  theme(plot.title = element_text(hjust = 0.5))

coverage_bootstrap_plot <-
  ggplot(plot_data, aes(x = n)) +
  geom_line(aes(y = Coverage_Bootstrap), color = "darkgreen") +
  geom_hline(yintercept = 0.95, linetype = "dashed") +
  labs(y = "Coverage", x = "Sample Size (n)", title = "Bootstrap CI") +
  ylim(0, 1) +
  theme(plot.title = element_text(hjust = 0.5))

coverage_percentile_plot <-
  ggplot(plot_data, aes(x = n)) +
  geom_line(aes(y = Coverage_Percentile), color = "orange") +
  geom_hline(yintercept = 0.95, linetype = "dashed") +
  labs(y = "Coverage", x = "Sample Size (n)", title = "Percentile CI") +
  ylim(0, 1) +
  theme(plot.title = element_text(hjust = 0.5))

library(gridExtra)
grid.arrange(
  bias_plot, variance_plot, mse_plot,
  coverage_normal_plot, coverage_bootstrap_plot, coverage_percentile_plot,
  ncol = 3, nrow = 2
)
```

    ii. 
Bias is around 0 for all sample sizes, indicating it's a unbiased estimator. When n increases, estimated variance and MSE decreases and approaches zero. The coverage for all three confidence intervals is close to 95%.

\newpage

## QUESTION 4: Randomization Test [12 points]

**(a)**

```{r}
library(ggplot2)
library(dplyr)
nba <- read.csv("NBA_2024_Shots.csv")
S1 <- subset(nba, nba$TEAM_NAME_ABBREV == nba$HOME_TEAM)  # Home team shots
S2 <- subset(nba, nba$TEAM_NAME_ABBREV == nba$AWAY_TEAM)  # Away team shots

# FG% for home and away
FG_home <- 100 * sum(S1$SHOT_MADE) / nrow(S1)
FG_away <- 100 * sum(S2$SHOT_MADE) / nrow(S2)

cat("FG% Home:", FG_home, "\n")
cat("FG% Away:", FG_away, "\n")

S1$Category <- "Home"
S2$Category <- "Away"
plot_data <- rbind(S1, S2)

# Data setup
made <- c(47.08, 47.79)       # FG% for Away and Home
missed <- 100 - made          # Missed shots percentage
data <- rbind(made, missed)   # Combine into a matrix

bar_colors <- c("forestgreen", "brown3") 
bar_labels <- c("Away", "Home")          

bar_heights <- barplot(data, beside = FALSE, 
                       col = bar_colors, border = "black", 
                       names.arg = bar_labels, 
                       ylim = c(0, 100), 
                       ylab = "Percentage of Shots", 
                       main = "Home vs. Away Shot Outcomes")

text(x = bar_heights, y = made + 5, 
     labels = paste0("FG% = ", round(made, 2), "%"), 
     col = "black", font = 2)

legend("topright", legend = c("Made", "Missed"), 
       fill = bar_colors, border = "black")

```

$\;$

**(b)**
If $S_1$ and $S_2$ are randomly drawn, they represent their corresponding populations $P_1$ and $P_2$. 
If $H_0$: $S_1$ and $S_2$ are drawn from the same population, this implies that $P_1=P_2$. 
Oppositely, if $H_A:P_1\neq P_2$, there'll be a difference between $S_1$ and $S_2$.
Thus, these statements are equivalent.

$\;$

**(c)**

    i. 
```{r}
obs_discrepancy <- abs(FG_home - FG_away)
obs_discrepancy
```
    ii. 
```{r}
set.seed(123)
M <- 1000  
discrepancies <- numeric(M)

S1_original <- subset(plot_data, Category == "Home")
S2_original <- subset(plot_data, Category == "Away")

mix_randomly <- function(pop1, pop2) {
  # population sizes
  n_pop1 <- nrow(pop1)
  n_pop2 <- nrow(pop2)
  # combined population
  full_pop <- rbind(pop1, pop2)
  # pick the random indices for pop1
  rid_pop1 <- sample(n_pop1 + n_pop2, n_pop1)
  # create the mixed populations
  new_pop1 <- full_pop[rid_pop1, ]
  new_pop2 <- full_pop[-rid_pop1, ]
  # output
  list(pop1 = new_pop1, pop2 = new_pop2)
}

for (i in 1:M) {
  # Mix the two populations
  mixed_populations <- mix_randomly(S1_original, S2_original)
  S1_shuffled <- mixed_populations$pop1
  S2_shuffled <- mixed_populations$pop2
  
  # Calculate FG% for mixed data
  FG_home_shuffled <- 100 * sum(S1_shuffled$SHOT_MADE) / nrow(S1_shuffled)
  FG_away_shuffled <- 100 * sum(S2_shuffled$SHOT_MADE) / nrow(S2_shuffled)
  
  # Calculate the discrepancy
  discrepancies[i] <- abs(FG_home_shuffled - FG_away_shuffled)
}

ggplot(data.frame(discrepancies), aes(x = discrepancies)) +
  geom_histogram(bins = 15, color = "black", fill = "lightblue") +
  geom_vline(xintercept = obs_discrepancy, color = "red", linetype = "dashed") +
  labs(title = "Randomization Test", x = "Discrepancy", y = "Frequency") +
  theme(plot.title = element_text(hjust = 0.5))

```
    iii. 
```{r}
p_value <- mean(discrepancies >= obs_discrepancy)
cat("P-value:", p_value, "\n")
```
    iv. 
p-val is 0.001. There is very strong evidence against $H_0$ based on the observed data, meaning that there's a significant difference between $FG\%_{home}$ and $FG\%_{away}$. This conclusion doesn't coincide with the conclusion that we can draw from part (a), as in part (a) the difference is not significant. Thus, even if the difference is statistically significant, it might not be practically meaningful.

\newpage


## QUESTION 5: Estimation via 2-D Smoothing [13 points] 

**(a)**
```{r fig.height=6, fig.width=5}
plot(nba$LOC_X, nba$LOC_Y, 
     col = ifelse(nba$SHOT_MADE, adjustcolor("forestgreen", 0.2), 
                                 adjustcolor("firebrick", 0.2)),
     pch = 16, cex = 0.5, 
     xlab = "Court Width", ylab = "Court Length",
     main = "Scatter Plot of NBA Shots",
     xlim = c(-30, 30), ylim = c(0, 95), asp = 1)

draw_court(add = TRUE)
```

$\;$

**(b)**
```{r fig.height=6, fig.width=5}
# Set up grid
n <- 50  # Number of grid divisions
x_seq <- seq(-25, 25, length.out = n)  # X coordinates from -25 to 25
y_seq <- seq(0, 94, length.out = n)     # Y coordinates from 0 to 94
theta_estimates <- matrix(NA, nrow = n, ncol = n)

# Calculate estimates
for (i in seq_along(x_seq)) {
  for (j in seq_along(y_seq)) {
    x0 <- x_seq[i]
    y0 <- y_seq[j]
    
    # Identify neighborhood within 10 feet
    neighborhood <- nba %>%
                    filter(LOC_X >= (x0 - 10) & LOC_X <= (x0 + 10) &
                           LOC_Y >= (y0 - 10) & LOC_Y <= (y0 + 10))

    if (nrow(neighborhood) > 0) {
      weights <- 1 / (1 + sqrt((neighborhood$LOC_X - x0)^2 + (neighborhood$LOC_Y - y0)^2))
      theta_estimates[i, j] <- sum(weights * neighborhood$SHOT_MADE) / sum(weights)
    } else {
      theta_estimates[i, j] <- NA  # Not enough data in this neighborhood
    }
  }
}

# Transpose the matrix and reverse the y-axis
corrected_theta_estimates <- apply(theta_estimates, 2, rev)

# Create the heatmap with corrected orientation
image(x_seq, y_seq, corrected_theta_estimates,
      col = colorRampPalette(c("firebrick", "forestgreen"))(100),
      xlab = "Court Width", ylab = "Court Length", 
      main = "Estimated Shooting Accuracy",
      axes = TRUE)

# Overlay the court
draw_court(add = TRUE)

# Add contours
contour(x_seq, y_seq, corrected_theta_estimates, add = TRUE, col = "blue")

```

$\;$

**(c)**
```{r}
# Function to calculate the estimate for bootstrapping
calculate_theta <- function(data, x0, y0) {
  neighborhood <- data %>%
                  filter(LOC_X >= (x0 - 10) & LOC_X <= (x0 + 10) & 
                         LOC_Y >= (y0 - 10) & LOC_Y <= (y0 + 10))

  if (nrow(neighborhood) > 0) {
    weights <- 1 / (1 + sqrt((neighborhood$LOC_X - x0) ^ 2 + 
                             (neighborhood$LOC_Y - y0) ^ 2))
    return(sum(weights * neighborhood$SHOT_MADE) / sum(weights))
  } else {
    return(NA)
  }
}

positions <- list(c(0, 16), c(0, 30), c(0, 47))
B <- 100
bootstrap_results <- list()

for (pos in positions) {
  x0 <- pos[1]
  y0 <- pos[2]
  
  estimates <- numeric(B)
  
  for (b in 1:B) {
    sample_data <- nba[sample(1:nrow(nba), replace = TRUE), ]
    estimates[b] <- calculate_theta(sample_data, x0, y0)
  }
  
  bootstrap_results[[paste0("Position: (", x0, ", ", y0, ")")]] <- 
    quantile(estimates, c(0.025, 0.975))
  cat("Position:", pos, "Estimate:", calculate_theta(nba, x0, y0), "CI:", 
      quantile(estimates, c(0.025, 0.975)), "\n")
}
```

\newpage

## APPENDIX A: Programming Guidelines

\newcommand{\tth}{\bm{\theta}}

#### Motivation


In Assignments 1 and 2, we introduced **good documentation** as the first fundamental pillar of good programming habits.  In this Assignment, we introduce the second fundamental pillar: **testing your code**.

The benefits of testing your code are largely self-evident: how else would you convince yourself -- or others -- that your code is correct?  

Testing your code can also be an important part of the debugging process.  Take for example Q2 on Assignment 2, in which we were asked to investigate the behavior of Gradient Descent and Newton-Raphson algorithms on the Beale function

$$
\rho(\theta_1, \theta_2) = (1.5 - \theta_1 + \theta_1 \theta_2)^2 + (2.25 - \theta_1 + \theta_1 \theta_2^2)^2 + (2.625 - \theta_1 + \theta_1 \theta_2^3)^2.
$$

Many students came to office hours wondering why these algorithms seemed to throw errors or return nonsense.  After much pain, it was typically discovered that the derivatives $\psi(\tth) = \grad \rho(\tth)$ and/or $\psi'(\tth)$ were implemented incorrectly.  This was much more difficult and time-consuming to track down from running the functions `gradientDescent()` and `NewtonRaphson()` with all their subfunctions and other moving pieces, rather than first testing that the derivatives are correct.  

Indeed, some would go as far as to say that testing your code extensively leads to better code design.  For example, `gradientDescent()` allows users to pass in their own `lineSearchFn` and `testConvergenceFn`.  However, an initial implementation did not provide this feature.  Rather, only `gridLineSearch()` and `testConvergence()` were provided as nested functions defined inside of `gradientDescent()`.  However, when we realized that it was extremely difficult to check that these nested functions were doing what they were supposed to, we added the feature that users could provide their own line search and convergence functions.  This easily allowed us to test our implementations of `gridLineSearch()` and `testConvergence()`, while also creating a more versatile and general-purpose implementation of `gradientDescent()`.  In a nutshell, if you are struggling to test a complicated function, then chances are you can refactor it into subfunctions which are easier to test, and this will likely improve the clarity, simplicity, and flexibility of your code.

A few more arguments for testing your code (in the context of larger and/or collaborative projects) are given [here](https://r-pkgs.org/testing-basics.html).

#### Informal Testing

The situation we described above is called a [unit test](https://en.wikipedia.org/wiki/Unit_testing), i.e., testing small components (or units) of a larger piece of code prior to testing all of it together.  Let's consider the example in more detail.  Specifically, suppose we wish to test our implementation of $\grad \rho(\tth) = (\pdv{\rho(\tth)}{\theta_1}, \pdv{\rho(\tth)}{\theta_2})$, which is

$$
\begin{aligned}
\pdv{\rho(\tth)}{\theta_1} & = 2\left(1.5-\theta_1+\theta_1\theta_2\right)(\theta_2-1) + 2\left(2.25-\theta_1+\theta_1\theta_2^2\right)\left(\theta_2^2-1\right)+2\left(2.625-\theta_1+\theta_1\theta_2^3\right)\left(\theta_2^3-1\right), \\
\pdv{\rho(\tth)}{\theta_2} & = 2\left(1.5-\theta_1+\theta_1\theta_2\right)(\theta_1) + 2\left(2.25-\theta_1+\theta_1\theta_2^2\right)\left(2\theta_1\theta_2\right)+2\left(2.625-\theta_1+\theta_1\theta_2^3\right)\left(3\theta_1\theta_2^2\right).
\end{aligned}
$$

An extremely common type of unit test is to check that two implementations of the same computation produce the same value.  For example:

- The simplified vs unsimplified implementation of a given formula.

- The result of a function optimizer algorithm (e.g., `gradientDescent()` gives the known minimum on a given problem like least-squares).

- That two optimization algorithms for the same function give the same answer (e.g., `gradientDescent()` vs `NewtonRaphson()`).

In order to check our implementation of $\grad \rho(\tth)$, we'll check it against a high-accuracy numerical approximation^[In a nutshell, this is done by computing $A_n = \frac{\rho(\theta_1 + \varepsilon_n, \theta_2) - \rho(\theta_1, \theta_2)}{\varepsilon_n}$ for a sequence $\varepsilon_1 > \varepsilon_2 > \cdots > \varepsilon_N$, then extrapolating the limit of $A_1, A_2, \ldots, A_N$ using a technique called [Richardson extrapolation](https://en.wikipedia.org/wiki/Richardson_extrapolation).  This limit is $\pdv{\rho(\tth)}{\theta_1}$.  A similar procedure is then used to obtain $\pdv{\rho(\tth)}{\theta_2}$.  The problem with this approach is not only that the extrapolation sometimes fails.  It is also that the number of evaluations of $\rho(\tth)$ scales with the number of elements of $\tth$.  Imagine how expensive this would be for $\tth \in \mathbb{R}^{100}$!] of the gradient provided by the R package [**numDeriv**](https://CRAN.R-project.org/package=numDeriv).  An *informal* unit test would then create functions to compute $\grad \rho(\tth)$ using both the analytic formula above and the numerical approximation with **numDeriv**, and check that they are sufficiently close on a few values:

```{r informaltest}
#' The Beale function itself: `rho(theta)`.
beale_rho <- function(theta) {
  term1 <- 1.5 - theta[1] + theta[1] * theta[2]
  term2 <- 2.25 - theta[1] + theta[1] * theta[2]^2
  term3 <- 2.625 - theta[1] + theta[1] * theta[2]^3
  term1^2 + term2^2 + term3^2
}

#' The gradient of the Beal function: `d/d theta rho(theta)`.
beale_gradient <- function(theta) {
  term1 <- 1.5 - theta[1] + theta[1] * theta[2]
  term2 <- 2.25 - theta[1] + theta[1] * theta[2]^2
  term3 <- 2.625 - theta[1] + theta[1] * theta[2]^3
  2 * c(
    # d rho / d theta1
    term1 * (-1 + theta[2]) +
      term2 * (-1 + theta[2]^2) +
      term3 * (-1 + theta[2]^3),
    # d rho / d theta2
    term1 * theta[1] +
      term2 * (2 * theta[1] * theta[2]) +
      term3 * (3 * theta[1] * theta[2]^2)
  )
}

#' Gradient of the Beale function approximated with numDeriv.
beale_gradient_numderiv <- function(theta) {
  numDeriv::grad(beale_rho, x = theta)
}

# check that the two computations of the gradient give the same value
compare_gradient <- function(theta) {
  rbind(
    analytic = beale_gradient(theta),
    numerical = beale_gradient_numderiv(theta)
  )
}

compare_gradient(theta = c(0, 0))
compare_gradient(theta = c(1, 0))
compare_gradient(theta = c(2, 0))
```

#### Formal Testing in R

The informal test above is the right place to start.  However there are a few conspicuous things that are lacking:

i. Suppose we want to refactor `beale_gradient()` to be more computationally efficient -- for example, by computing $\theta_2^k$ and $\theta_1 \theta_2^k$ only once for each $k = 1, 2, 3$.  To test this new implementation of `beale_gradient()` we essentially have to manually rerun the lines above and look at several lines of output.  This quickly gets tedious if we end up modifying `beale_gradient()` several times.  Thus, the unit tests should be **automated**.

ii. We used whole numbers for testing.  However, whole numbers are much more likely to "double cancel" out numerical errors than random numbers do.  Therefore, unit tests should employ **random numerical inputs** whenever possible.

iii. We tried a few simple inputs for testing.  However, in all of our tests we had $\theta_2 = 0$.  Good unit tests should **systematically cover as many cases** as possible, including edge cases.  In this setting, let's suppose that "cases" refer to each of $\theta_1$ and $\theta_2$ being negative, positive, or zero.

With these points in mind, let's convert the informal unit tests above into a formal unit test.  For this we'll use the R [**testthat**](https://testthat.r-lib.org/).  The most important functions in this package are `test_that()` and `expect_equal()`, of which the usage is illustrated below.

```{r formaltest}
library(testthat) # load the testthat library -- you need to install it first!

# data.frame containing all "cases" to test.
# here, it's just which elements of theta are equal to zero. 
test_descr <- expand.grid(
  theta1_sign = c(-1, 1, 0),
  theta2_sign = c(-1, 1, 0)
)
test_descr

# now let's loop through each case in a test_that environment
test_that("`beale_grad()` gives same result as numDeriv", {
  n_rep <- 10 # the number of times to repeat each test
  for(i in 1:nrow(test_descr)) {
    # info for the given test
    theta1_sign <- test_descr$theta1_sign[i]
    theta2_sign <- test_descr$theta2_sign[i]
    for(j in 1:n_rep) {
      # randomly generate theta for the given case
      theta <- runif(2, 0, 5)
      theta[1] <- theta[1] * theta1_sign
      theta[2] <- theta[2] * theta2_sign
      # analytic gradient
      grad_analytic <- beale_gradient(theta)
      grad_numderiv <- beale_gradient_numderiv(theta)
      expect_equal(
        grad_numderiv, grad_analytic,
        tolerance = 1e-8 # for roundoff errors
      )
    }
  }
})
```


\newpage

## APPENDIX B: Helpful Functions

```{r}
#' Draw an NBA basketball court to scale.
#'
#' @param add  A boolean input controlling whether the court is added to an existing plot 
#'             (add=TRUE) or created from scratch (add=FALSE).
#' @param lwd  A numeric input dictating the widths of the lines that define the court.
#' @param col  The colour of the lines that define the court.
#' @param axes A boolean input controlling whether the axis ticks are shown.
#' 
#' @return An NBA basketball court drawn as a new plot or added to an existing plot.
draw_court <- function(add = FALSE, lwd = 1, col = "black", axes = FALSE) {
  if(!add) {
    plot(
      x = 0, y = 0,
      col = "white", asp = 1,
      xlim = c(-30,30), xaxt = "n", xlab = "",
      ylim = c(0,95), yaxt = "n", ylab = ""
    )
    if(axes) {
      axis(side = 1, cex.axis = 0.75)
      axis(side = 2, cex.axis = 0.75)
    }
  }
  segments(x0 = -25, y0 = 0, x1 = 25, y1 = 0, lwd = lwd, col = col)
  segments(x0 = -25, y0 = 94, x1 = 25, y1 = 94, lwd = lwd, col = col)
  segments(x0 = -25, y0 = 0, x1 = -25, y1 = 94, lwd = lwd, col = col)
  segments(x0 = 25, y0 = 0, x1 = 25, y1 = 94, lwd = lwd, col = col)
  segments(x0 = -25, y0 = 47, x1 = 25, y1 = 47, lwd = lwd, col = col)
  lines(x = seq(-6,6,0.01), y = sqrt(36-seq(-6,6,0.01)^2)+47, lwd = lwd, col = col)
  lines(x = seq(-6,6,0.01), y = -sqrt(36-seq(-6,6,0.01)^2)+47, lwd = lwd, col = col)
  lines(x = seq(-22,22,0.01), y = sqrt(23.9^2-seq(-22,22,0.01)^2)+4.7, lwd = lwd, col = col)
  segments(x0 = -22, y0 = 0, x1 = -22, y1 = 14, lwd = lwd, col = col)
  segments(x0 = 22, y0 = 0, x1 = 22, y1 = 14,lwd = lwd, col = col)
  lines(x = seq(-22,22,0.01), y = -sqrt(23.9^2-seq(-22,22,0.01)^2)+89.3, lwd = lwd, col = col)
  segments(x0 = -22, y0 = 80, x1 = -22, y1 = 94, lwd = lwd, col = col)
  segments(x0 = 22, y0 = 80, x1 = 22, y1 = 94, lwd = lwd, col = col)
  lines(x = seq(-6,6,0.01), y = sqrt(36-seq(-6,6,0.01)^2)+15, lwd = lwd, col = col)
  lines(x = seq(-6,6,0.01), y = -sqrt(36-seq(-6,6,0.01)^2)+15, lwd = lwd, lty = 3, col = col)
  segments(x0 = -8, y0 = 0, x1 = -8, y1 = 15, lwd = lwd, col = col)
  segments(x0 = 8, y0 = 0, x1 = 8, y1 = 15, lwd = lwd, col = col)
  segments(x0 = -8, y0 = 15, x1 = 8, y1 = 15, lwd = lwd, col = col)
  lines(x = seq(-6,6,0.01), y = -sqrt(36-seq(-6,6,0.01)^2)+79, lwd = lwd, col = col)
  lines(x = seq(-6,6,0.01), y = sqrt(36-seq(-6,6,0.01)^2)+79, lwd = lwd, lty = 3, col = col)
  segments(x0 = -8, y0 = 94, x1 = -8, y1 = 79, lwd = lwd, col = col)
  segments(x0 = 8, y0 = 94, x1 = 8, y1 = 79, lwd = lwd, col = col)
  segments(x0 = -8, y0 = 79, x1 = 8, y1 = 79, lwd = lwd, col = col)
  points(x = 0, y = 4, pch = 16, col = col)
  segments(x0 = -2, y0 = 3, x1 = 2, y1 = 3, lwd = lwd, col = col)
  points(x = 0, y = 90, pch = 16, col = col)
  segments(x0 = -2, y0 = 91, x1 = 2, y1 = 91, lwd = lwd, col = col)
}
```

\newpage

```{r}
#' Horvitz-Thompson variance estimate.
#'
#' @param t_u The vector of `n` values of `t_u` for each unit in the sample.
#' @param pi_u The vector of `n` inclusion probabilities for each unit in the
#'   sample.
#' @param pi_uv A matrix of `n x n` joint inclusion probabilities for every pair 
#'   of the units in the sample.
#'
#' @return The Horvitz-Thompson variance estimate.
est_var_ht <- function(t_u, pi_u, pi_uv) {
  delta <- pi_uv - outer(pi_u, pi_u)
  var_hat <- sum((delta/pi_uv) * outer(t_u/pi_u, t_u/pi_u))
  return(var_hat)
}
```

```{r}
#' Randomly mix two populations.
#'
#' @param pop1 A data.frame containing population 1.
#' @param pop2 A data.frame containing population 2.
#'
#' @return A list with elements `pop1` and `pop2`, where each is a data.frame
#'   of the same size as the corresponding original input, but with units (rows)
#'   randomly mixed between the two populations.
#'
#' @details `pop1` and `pop2` must have the same column names and dimensions,
#'   i.e., one should be able to perform `rbind(pop1, pop2)` without throwing
#'   an error.
mix_randomly <- function(pop1, pop2) {
  # population sizes
  n_pop1 <- nrow(pop1)
  n_pop2 <- nrow(pop2)
  # combined population
  full_pop <- rbind(pop1, pop2)
  # pick the random indices for pop1
  rid_pop1 <- sample(n_pop1 + n_pop2, n_pop1)
  # create the mixed populations
  new_pop1 <- full_pop[rid_pop1, ]
  new_pop2 <- full_pop[-rid_pop1, ]
  # output
  list(pop1 = new_pop1, pop2 = new_pop2)
}
```

\newpage

```{r getjointinclusion}
#' Compute the joint inclusion probabilities for an unequally weighted sampling
#'   without replacement scheme.
#'
#' @param n Size of sample to draw.
#' @param prob Vector of `N` probabilities (between 0 and 1).
#'
#' @return `N x N` matrix of joint inclusion probabilities.
#'
#' @details The sampling scheme is that employed by
#'   ```
#'   sample(N, size = n, replace = TRUE, prob = prob)
#'   ```
#'   That is:
#' 
#'   - The first unit is drawn from `1:N` with probability `prob / sum(prob)`.
#'   - If the first unit is `i`, the second unit is drawn from `(1:N)[-i]`
#'     with probability `prob[-i] / sum(prob[-i])`.
#'   - If the first two units are `i` and `j`, then the third unit is drawn from
#'   `(1:N)[-c(i,j)]` with probability `prob[-c(i,j)]/sum(prob[-c(i,j)])`,
#'     and so on.
get_joint_inclusion <- function(size, prob) {
  # all unordered samples
  N <- length(prob)
  all_samples <- combn(N, size)
  n_samples <- ncol(all_samples)
  # all permutations to get ordered samples
  all_perms <- as.matrix(perm(1:size))
  # compute probability of each unordered sample
  all_prob <- rep(0, n_samples)
  for(i in 1:ncol(all_perms)) {
    cum_prob <- 1 # cumulative probability of ordered sample
    rem_prob <- rep(sum(prob), n_samples) # remaining probabilty
    for(j in 1:size) {
      pr <- prob[all_samples[all_perms[j,i],]]
      cum_prob <- cum_prob * (pr / rem_prob)
      rem_prob <- rem_prob - pr
    }
    all_prob <- all_prob + cum_prob
  }
  # compute the joint inclusion matrix
  pi_mat <- matrix(0, N, N)
  for(i in 1:N) {
    for(j in 1:N) {
      # which samples contain (i, j)
      id <- (i == all_samples) + (j == all_samples)
      id <- colSums(id) == 2
      pi_mat[i, j] <- sum(all_prob[id])
    }
  }
  pi_mat
}
```

\newpage

```{r perm}
#' Compute all permutations of the input.
#'
#' @param x A numeric vector of length `n`.
#'
#' @return A matrix of size `n x factorial(n)` containing all permutations of `x`.
perm <- function(x) {
  if (length(x) == 1) {
    return(x)
  } else {
    res <- matrix(ncol = 0, nrow = length(x))
    for (i in seq_along(x)) {
      res <- cbind(res, rbind(x[i], Recall(x[-i])))
    }
    return(res)
  }
}
```
