---
title: 'STAT 341: Assignment 2'
subtitle: 'DUE: Tuesday November 5, 2024 by 5:00pm ET'
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
urlcolor: blue
header-include: \usepackage{amsmath}
---
```{r setup, include = FALSE}
library(knitr)
library(ggplot2)
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  tidy = "styler"
)
```

$\;$
$\;$
$\;$
$\;$

## NOTES

Your assignment must be submitted by the due date listed at the top of this document, and it must be submitted electronically in .pdf format via Crowdmark. This means that your responses for different questions should begin on separate pages of your .pdf file. Note that your .pdf solution file must have been generated by R Markdown. Additionally:

* For mathematical questions: your solutions must be produced by LaTeX (from within R Markdown). Neither screenshots nor scanned/photographed handwritten solutions will be accepted -- these will receive zero points.

* For computational questions: R code should always be included in your solution (via code chunks in R Markdown). If code is required and you provide none, you will receive zero points.
   + **Exception:** any functions defined in the lecture notes can be loaded using `echo=FALSE` but any other code chunks should have `echo=TRUE`. e.g., the code chunk loading `gradientDescent()` can use `echo=FALSE` but chunks that call `gradientDescent()` should have `echo=TRUE`.
   
* For interpretation questions: plain text (within R Markdown) is required. Text responses embedded as comments within code chunks will not be accepted.

Organization and comprehensibility is part of a full solution. Consequently, points will be deducted for solutions that are not organized and incomprehensible. Furthermore, if you submit your assignment to Crowdmark, but you do so incorrectly in any way (e.g., you upload your Question 2 solution in the Question 1 box), you will receive a 5% deduction (i.e., 5% of the assignment’s point total will be deducted from your point total).

\newpage

## QUESTION 1: Coding Best Practices [2 points]

This isn't so much an independent question as much as it's a motivator to use good coding practices when creating your solutions to this assignment. Read the **new** "Programming Guidelines" in Appendix A at the end of this document and do your best to adhere to these guidelines when preparing your solutions. You will be graded as follows:

* 2 points: good effort was taken to follow the stated programming guidelines
* 1 point: some effort was taken to follow the stated programming guidelines
* 0 points: no effort was taken to follow the stated programming guidelines

\newpage


## QUESTION 2: The Beale Function [36 points]

**(a)** 
$$ 
\frac{\partial \rho}{\partial \theta_1} 
= 2(1.5 - \theta_1 + \theta_1 \theta_2)(-1 + \theta_2) + 
2(2.25 - \theta_1 + \theta_1 \theta_2^2)(-1 + \theta_2^2) + 
2(2.625 - \theta_1 + \theta_1 \theta_2^3)(-1 + \theta_2^3) \\
$$
$$
\frac{\partial \rho}{\partial \theta_2} = 
2(1.5 - \theta_1 + \theta_1 \theta_2) \cdot \theta_1 + 
2(2.25 - \theta_1 + \theta_1 \theta_2^2) \cdot 2 \theta_1 \theta_2 + 
2(2.625 - \theta_1 + \theta_1 \theta_2^3) \cdot 3 \theta_1 \theta_2^2 \\
$$
$$
\boldsymbol{g} =  \nabla\rho(\boldsymbol{\theta}) = \begin{bmatrix}
2(1.5 - \theta_1 + \theta_1 \theta_2)(-1 + \theta_2) + 
2(2.25 - \theta_1 + \theta_1 \theta_2^2)(-1 + \theta_2^2) + 
2(2.625 - \theta_1 + \theta_1 \theta_2^3)(-1 + \theta_2^3)\\
2(1.5 - \theta_1 + \theta_1 \theta_2) \cdot \theta_1 + 
2(2.25 - \theta_1 + \theta_1 \theta_2^2) \cdot 2 \theta_1 \theta_2 + 
2(2.625 - \theta_1 + \theta_1 \theta_2^3) \cdot 3 \theta_1 \theta_2^2
\end{bmatrix}
$$

**(b)**
```{r}
beale_rho <- function(theta) {
  theta1 <- theta[1]
  theta2 <- theta[2]
  
  term1 <- (1.5 - theta1 + theta1 * theta2)^2
  term2 <- (2.25 - theta1 + theta1 * theta2^2)^2
  term3 <- (2.625 - theta1 + theta1 * theta2^3)^2
  
  return(term1 + term2 + term3)
}

beale_gradient <- function(theta) {
  theta1 <- theta[1]
  theta2 <- theta[2]
  
  grad_theta1 <- 2 * (1.5 - theta1 + theta1 * theta2) * (-1 + theta2) +
                 2 * (2.25 - theta1 + theta1 * theta2^2) * (-1 + theta2^2) +
                 2 * (2.625 - theta1 + theta1 * theta2^3) * (-1 + theta2^3)
  
  grad_theta2 <- 2 * (1.5 - theta1 + theta1 * theta2) * theta1 +
                 2 * (2.25 - theta1 + theta1 * theta2^2) * 2 * theta1 * theta2 +
                 2 * (2.625 - theta1 + theta1 * theta2^3) * 3 * theta1 * theta2^2
  
  return(c(grad_theta1, grad_theta2))
}
```


**(c)**
```{r}
gradientDescent <- function(theta = 0, rhoFn, gradientFn, lineSearchFn,
  testConvergenceFn, maxIterations = 100, tolerance = 1e-06, relative = FALSE, 
  lambdaStepsize = 0.01, lambdaMax = 0.5) {
    converged <- FALSE
    i <- 0

    while (!converged & i <= maxIterations) {
      g <- gradientFn(theta) ## gradient
      glength <- sqrt(sum(g^2)) ## gradient direction
      if (glength > 0)
        d <- g/glength

      lambda <- lineSearchFn(theta, rhoFn, d, lambdaStepsize = lambdaStepsize,
                             lambdaMax = lambdaMax)

      thetaNew <- theta - lambda * d
      converged <- testConvergenceFn(thetaNew, theta, tolerance = tolerance, 
                                     relative = relative)
      theta <- thetaNew
      i <- i + 1
    }
    ## Return last value and whether converged or not
    list(theta = theta, converged = converged, iteration = i, 
         fnValue = rhoFn(theta))
}

### line searching could be done as a simple grid search
gridLineSearch <- function(theta, rhoFn, d, lambdaStepsize = 0.01, lambdaMax = 1) {
    ## grid of lambda values to search
    lambdas <- seq(from = 0, by = lambdaStepsize, to = lambdaMax)
    ## line search
    rhoVals <- sapply(lambdas, function(lambda) {
        rhoFn(theta - lambda * d)
    })
    ## Return the lambda that gave the minimum
    lambdas[which.min(rhoVals)]
}

### Where testCovergence might be (relative or absolute)
testConvergence <- function(thetaNew, thetaOld, tolerance = 1e-10, relative = FALSE) {
    sum(abs(thetaNew - thetaOld)) < if (relative)
        tolerance * sum(abs(thetaOld)) else tolerance
}

gradientDescent(theta = c(3, 3), rhoFn = beale_rho, gradientFn = beale_gradient, 
                lineSearchFn = gridLineSearch, lambdaStepsize = 1e-03, 
                testConvergenceFn = testConvergence, maxIterations = 500)
gradientDescent(theta = c(3, -3), rhoFn = beale_rho, gradientFn = beale_gradient, 
                lineSearchFn = gridLineSearch, lambdaStepsize = 1e-03, 
                testConvergenceFn = testConvergence, maxIterations = 500)
gradientDescent(theta = c(-3, -3), rhoFn = beale_rho, gradientFn = beale_gradient, 
                lineSearchFn = gridLineSearch, lambdaStepsize = 1e-03, 
                testConvergenceFn = testConvergence, maxIterations = 500)
gradientDescent(theta = c(-3, 3), rhoFn = beale_rho, gradientFn = beale_gradient, 
                lineSearchFn = gridLineSearch, lambdaStepsize = 1e-03, 
                testConvergenceFn = testConvergence, maxIterations = 500)

```

**(d)**
```{r}
# Modified gradientDescent with iterations
gradientDescent <- function(theta = 0, rhoFn, gradientFn, lineSearchFn,
  testConvergenceFn, maxIterations = 100, tolerance = 1e-06, 
  relative = FALSE, lambdaStepsize = 0.01, lambdaMax = 0.5) {
  
  converged <- FALSE
  i <- 0
  iterates <- list(theta) # Add iterates
  
  while (!converged & i <= maxIterations) {
    g <- gradientFn(theta)
    glength <- sqrt(sum(g^2)) 
    if (glength > 0) 
      d <- g / glength  
      
    lambda <- lineSearchFn(theta, rhoFn, d, lambdaStepsize = lambdaStepsize,
                             lambdaMax = lambdaMax)

    thetaNew <- theta - lambda * d

    iterates[[i + 1]] <- thetaNew # Update iterates
    converged <- testConvergenceFn(thetaNew, theta, tolerance = tolerance, 
                                     relative = relative)
    theta <- thetaNew
    i <- i + 1
  }
  list(theta = theta, converged = converged, iteration = i, 
       fnValue = rhoFn(theta), iterates = iterates)
}

# Define the four starting points
starting_points <- list(
  c(3, 3),
  c(3, -3),
  c(-3, -3),
  c(-3, 3)
)

results <- lapply(starting_points, function(start) {
  gradientDescent(
    theta = start,
    rhoFn = beale_rho,
    gradientFn = beale_gradient,
    lineSearchFn = gridLineSearch,
    testConvergenceFn = testConvergence,
    maxIterations = 500,
    lambdaStepsize = 1e-3
  )
})

# Generate a grid of values for theta1 and theta2 for plotting
theta1_vals <- seq(-5, 5, length.out = 100)
theta2_vals <- seq(-5, 5, length.out = 100)
grid <- expand.grid(theta1 = theta1_vals, theta2 = theta2_vals)
grid$rho <- apply(grid, 1, function(theta) beale_rho(c(theta[1], theta[2])))

# The contour plot
plot(grid$theta1, grid$theta2, type = "n", 
     xlab = expression(theta[1]), 
     ylab = expression(theta[2]), 
     main = "Beale Function with Gradient Descent Paths")
contour(unique(grid$theta1), unique(grid$theta2), matrix(grid$rho, nrow = 100), 
        levels = 10^(0:5), col = "darkgrey", add = TRUE)

# Plot global minimum
points(3, 0.5, pch = 18, col = "black", cex = 1.5)

# Add paths with segments
colors <- c("blue", "red", "green", "purple")
for (i in 1:length(results)) {
  path <- results[[i]]$iterates
  
  for (j in 1:(length(path) - 1)) {
    segments(x0 = path[[j]][1], y0 = path[[j]][2], 
             x1 = path[[j + 1]][1], y1 = path[[j + 1]][2], 
             col = colors[i])
  }
  # Mark the starting and ending points of each path
  points(path[[1]][1], path[[1]][2], col = colors[i], pch = 16)
  points(path[[length(path)]][1], path[[length(path)]][2], 
         col = colors[i], pch = 18)
}
```

**(e)** Observe that different starting points have different paths and different convergence result. Starting values can impact the convergence path and the convergence result, as the optimization algorithm may converge to local minima rather than the global minimum.

**(f)**
$$
\psi(\theta) = \nabla \rho(\theta) = \begin{bmatrix} 
\frac{\partial \rho}{\partial \theta_1} \\ 
\frac{\partial \rho}{\partial \theta_2} \end{bmatrix}, where
$$
$$
\frac{\partial \rho}{\partial \theta_1} = 
2(1.5 - \theta_1 + \theta_1 \theta_2)(-1 + \theta_2) + 
2(2.25 - \theta_1 + \theta_1 \theta_2^2)(-1 + \theta_2^2) + 
2(2.625 - \theta_1 + \theta_1 \theta_2^3)(-1 + \theta_2^3)
$$
$$
\frac{\partial \rho}{\partial \theta_2} = 
2(1.5 - \theta_1 + \theta_1 \theta_2) \cdot \theta_1 + 
2(2.25 - \theta_1 + \theta_1 \theta_2^2) \cdot 2 \theta_1 \theta_2 + 
2(2.625 - \theta_1 + \theta_1 \theta_2^3) \cdot 3 \theta_1 \theta_2^2
$$
$$
\psi'(\theta) = \begin{bmatrix} 
\frac{\partial^2 \rho}{\partial \theta_1^2} & 
\frac{\partial^2 \rho}{\partial \theta_1 \partial \theta_2} \\ 
\frac{\partial^2 \rho}{\partial \theta_2 \partial \theta_1} & 
\frac{\partial^2 \rho}{\partial \theta_2^2} \end{bmatrix}, where
$$
$$
\frac{\partial^2 \rho}{\partial \theta_1^2} = 
2(-1 + \theta_2)^2 + 
2(-1 + \theta_2^2)^2 + 
2(-1 + \theta_2^3)^2
$$
$$
\frac{\partial^2 \rho}{\partial \theta_1 \partial \theta_2} =
\frac{\partial^2 \rho}{\partial \theta_2 \partial \theta_1} =
2(1.5 - 2\theta_1 + 2\theta_1 \theta_2) + 
4(2.25\theta_2 - 2\theta_1\theta_2 + 2\theta_1 \theta_2^3) + 
6(2.625\theta_2^2 - 2\theta_1\theta_2^2 + 2\theta_1\theta_2^5)
$$
$$
\frac{\partial^2 \rho}{\partial \theta_2^2} =
2\theta_1^2 + 8\theta_1^2\theta_2^2 + 
4\theta_1(2.25 - \theta_1 + \theta_1 \theta_2^2) + 
18\theta_1^2\theta_2^4 +
12\theta_1\theta_2(2.625 - \theta_1 + \theta_1 \theta_2^3)
$$

```{r}
beale_psi <- function(theta) {
  beale_gradient(theta = theta) # beale_psi is the same as beale_gradient
}

beale_psi_prime <- function(theta) {
  theta1 <- theta[1]
  theta2 <- theta[2]
  
  # Second partial derivative with respect to theta1
  hess_theta1_theta1 <- 2 * ((-1 + theta2)^2) +
                        2 * ((-1 + theta2^2)^2) +
                        2 * ((-1 + theta2^3)^2)
  
  # Mixed partial derivative
  hess_theta1_theta2 <- 2 * (1.5 - 2 * theta1 + 2 * theta1 * theta2) +
    4 * theta2 * (2.25 - 2 * theta1 + 2 * theta1 * theta2^2) +
    6 * theta2^2 * (2.625 - 2 * theta1 + 2 * theta1 * theta2^3)
  
  # Second partial derivative with respect to theta2
  hess_theta2_theta2 <- 2 * theta1^2 +
                        8 * theta1^2 * theta2^2 + 
                        4 * theta1 * (2.25 - theta1 + theta1 * theta2^2) +
                        18 * theta1^2 * theta2^4 +
                        12 * theta1 * theta2 * (2.625 - theta1 + 
                                                theta1 * theta2^3)
  
  # Construct the Jacobian matrix
  jacobian_matrix <- matrix(c(hess_theta1_theta1, hess_theta1_theta2,
                              hess_theta1_theta2, hess_theta2_theta2), 
                            nrow = 2, ncol = 2, byrow = TRUE)
  
  return(jacobian_matrix)
}
```

**(g)**
```{r}
NewtonRaphson <- function(theta, psiFn, psiPrimeFn, dim, 
  testConvergenceFn = testConvergence, maxIterations = 100, 
  tolerance = 1e-06, relative = FALSE) {
    if (missing(theta)) {
        ## need to figure out the dimensionality
        if (missing(dim)) {
            dim <- length(psiFn())
        }
        theta <- rep(0, dim)
    }
    converged <- FALSE
    i <- 0
    while (!converged & i <= maxIterations) {
        thetaNew <- theta - solve(psiPrimeFn(theta), psiFn(theta))
        converged <- testConvergenceFn(thetaNew, theta, tolerance = tolerance,
            relative = relative)

        theta <- thetaNew
        i <- i + 1
    }
    ## Return last value and whether converged or not
    list(theta = theta, converged = converged, 
         iteration = i, fnValue = psiFn(theta))
}

NewtonRaphson(c(3,3), psiFn = beale_psi, psiPrimeFn = beale_psi_prime, 
              maxIterations = 500)
NewtonRaphson(c(3,-3), psiFn = beale_psi, psiPrimeFn = beale_psi_prime, 
              maxIterations = 500)
NewtonRaphson(c(-3,-3), psiFn = beale_psi, psiPrimeFn = beale_psi_prime, 
              maxIterations = 500)
NewtonRaphson(c(-3,3), psiFn = beale_psi, psiPrimeFn = beale_psi_prime, 
              maxIterations = 500)
```
**(h)**
```{r}
NewtonRaphson <- function(theta, psiFn, psiPrimeFn, dim, 
  testConvergenceFn = testConvergence, maxIterations = 100, 
  tolerance = 1e-06, relative = FALSE) {
    if (missing(theta)) {
        # Determine dimensionality if theta is missing
        if (missing(dim)) {
            dim <- length(psiFn())
        }
        theta <- rep(0, dim)
    }
    converged <- FALSE
    i <- 0
    iterates <- list(theta)  # Initialize list to store each iterate
    
    while (!converged & i <= maxIterations) {
        # Newton-Raphson update
        thetaNew <- theta - solve(psiPrimeFn(theta), psiFn(theta))
        # Check for convergence
        converged <- testConvergenceFn(thetaNew, theta, tolerance = tolerance, 
                                       relative = relative)
        # Store the new iterate
        iterates[[i + 1]] <- thetaNew
        # Update theta for the next iteration
        theta <- thetaNew
        i <- i + 1
    }
    list(theta = theta, converged = converged, iteration = i, 
         fnValue = psiFn(theta), iterates = iterates)
}

results <- lapply(starting_points, function(start) {
  NewtonRaphson(
    theta = start,
    psiFn = beale_psi,
    psiPrimeFn = beale_psi_prime,
    testConvergenceFn = testConvergence,
    maxIterations = 500
  )
})

# Generate a grid of values for theta1 and theta2 for plotting
theta1_vals <- seq(-5, 5, length.out = 100)
theta2_vals <- seq(-5, 5, length.out = 100)
grid <- expand.grid(theta1 = theta1_vals, theta2 = theta2_vals)
grid$rho <- apply(grid, 1, function(theta) beale_rho(c(theta[1], theta[2])))

# The contour plot
plot(grid$theta1, grid$theta2, type = "n", 
     xlab = expression(theta[1]), 
     ylab = expression(theta[2]), 
     main = "Beale Function with Newton-Raphson Paths")
contour(unique(grid$theta1), unique(grid$theta2), matrix(grid$rho, nrow = 100), 
        levels = 10^(0:5), col = "darkgrey", add = TRUE)

# Plot global minimum
points(3, 0.5, pch = 18, col = "black", cex = 1.5)

# Add paths with segments
colors <- c("blue", "red", "green", "purple")
for (i in 1:length(results)) {
  path <- results[[i]]$iterates
  
  for (j in 1:(length(path) - 1)) {
    segments(x0 = path[[j]][1], y0 = path[[j]][2], 
             x1 = path[[j + 1]][1], y1 = path[[j + 1]][2], 
             col = colors[i])
  }
  points(path[[1]][1], path[[1]][2], col = colors[i], pch = 16)
  points(path[[length(path)]][1], path[[length(path)]][2], 
         col = colors[i], pch = 18)
}
```


**(i)**
No. The Newton-Raphson method can be sensitive to the choice of starting points and may fail to converge in non-convex, as it uses the curvature information that can lead it to local minima. In contrast, gradient descent often finds a more stable path by following the steepest descent. This behavior can make gradient descent more reliable on non-convex surfaces like the Beale function.

\newpage

## QUESTION 3: Stochastic Gradient Descent [15 points]

**(a)**
```{r}
## i
dummy_objective <- function(theta) {return(NULL)}

    
## ii
dummy_convergence_test <- function(thetaNew, thetaOld, tolerance = 1e-10, 
                                   relative = FALSE) {return(FALSE)}


## iii
decay_learning_rate <- function(theta, rhoFn, d, lambdaStepsize = 0.01, 
                                lambdaMax = 1, iteration = 1){
  return(10/(1 + iteration))
  }
    
gradientDescent <- function(theta = 0, rhoFn, gradientFn, lineSearchFn,
  testConvergenceFn, maxIterations = 100, tolerance = 1e-06, 
  relative = FALSE, lambdaStepsize = 0.01, lambdaMax = 0.5) {
    # Creating the path of results
    Path = matrix(NA, nrow = maxIterations + 1, ncol = length(theta))
    Path[1,] = theta
  
    converged <- FALSE
    i <- 0

    while (!converged & i <= maxIterations) {
      g <- gradientFn(theta) 
      glength <- sqrt(sum(g^2)) 
      if (glength > 0)
        d <- g/glength
      lambda <- lineSearchFn(theta, rhoFn, d, lambdaStepsize = lambdaStepsize,
                             lambdaMax = lambdaMax, iteration = i) 
      thetaNew <- theta - lambda * d
      converged <- testConvergenceFn(thetaNew, theta, tolerance = tolerance, 
                                     relative = relative)
      theta <- thetaNew
      i <- i + 1 
      if (i <= maxIterations){
      Path[(i+1),] = theta
      }
    }
    list(theta = theta, converged = converged, iteration = i, 
         fnValue = rhoFn(theta), Path = Path)
}


## iv

#' Create a Stochastic Gradient Function for Least Squares Regression
#'
#' @param x A numeric response vector of length 'N'.
#' @param y A numeric covariate vector of length 'N'.
#' @param n An integer specifying the size of the random subset of data points 
#'  to use in each gradient calculation.
#'
#' @return A vector of length 'p+1' specifying the coefficients of the line
#'  of best fit.
#'
#' @details Let
#'   ```
#'   y_hat(x, theta) = theta[1] + theta[2] * (x - xbar)
#'   ```
#'   denote the model where `theta[1]` represents the intercept (alpha) and
#'   `theta[2]` represents the slope (beta) of a simple linear regression. Here,
#'   `xbar` is the mean of the predictor variable `x`.
#'   
#'   The objective is to find the parameter vector `theta` that minimizes the
#'   least squares criterion over the entire dataset, i.e.,
#'   ```
#'   theta_hat = argmin_theta sum((y - y_hat(x, theta))^2).
#'   ```
#'   However, instead of using the full dataset, we approximate the gradient of
#'   this criterion by selecting a random subset of `n` data points in each 
#'   iteration.
#'   
#'   Let `subset` be a random sample of size `n` from the indices of `x` and `y`.
#'   The stochastic gradient with respect to `theta` for this subset is given by
#'   ```
#'   gradient(theta) = -2 * [
#'       sum(y_subset - theta[1] - theta[2] * (x_subset - xbar)),
#'       sum((y_subset - theta[1] - theta[2] * (x_subset - xbar)) * (x_subset - xbar))
#'   ]
#'   ```
#'   where `x_subset` and `y_subset` are the values of `x` and `y` for the sampled
#'   indices. This gradient is used in stochastic gradient descent to iteratively
#'   update `theta` toward the least squares solution.
create_least_squares_stochastic_gradient <- function(x, y, n) {
  xbar <- mean(x)
  ybar <- mean(y)
    
  function(theta) {
    selection <- sample(x = length(x), size = n)
    x <- x[selection]
    y <- y[selection]
    alpha <- theta[1]
    beta <- theta[2] 
    
    -2 * c(mean(y - alpha - beta * (x - xbar)), 
           mean((y - alpha - beta * (x - xbar)) * (x - xbar)))
  }
}
```

**(b)** 
```{r, fig.height = 8}
waldo <- read.csv("waldo.csv", header = TRUE) 
theta_init <- c(1,1)
x <- waldo$X
y <- waldo$Y

result <- gradientDescent(theta = theta_init, rhoFn = dummy_objective,
  gradientFn = create_least_squares_stochastic_gradient(x,y,20),
  lineSearchFn = decay_learning_rate,
  testConvergenceFn = dummy_convergence_test,
  maxIterations = 5000)

theta_pop <- lm(Y ~ I(X - mean(X)), data = waldo)

true_alpha <- coef(theta_pop)[1]
true_beta <- coef(theta_pop)[2]

par(mfrow = c(2, 1))

# plot alpha
plot(x = 0:5000, y = result$Path[,1], type = "l", col = "blue", 
     main = "SGD plot of alpha", xlab = "Iterations", ylab = "Estimate alpha")
abline(h = true_alpha, col = "red", lwd = 2, lty = 2)

# plot beta
plot(x = 0:5000, y = result$Path[,2], type = "l", col = "blue", 
     main = "SGD plot of beta", xlab = "Iterations", ylab = "Estimate beta") 
abline(h = true_beta, col = "red", lwd = 2, lty = 2)
```

**(c)**
```{r, fig.height = 8}
theta_init_2 <- c(10,10)

result_2 <- gradientDescent(theta = theta_init_2, rhoFn = dummy_objective,
  gradientFn = create_least_squares_stochastic_gradient(x,y,5),
  lineSearchFn = decay_learning_rate,
  testConvergenceFn = dummy_convergence_test,
  maxIterations = 5000)

par(mfrow = c(2, 1))

# plot alpha
plot(x = 0:5000, y = result_2$Path[,1], type = "l", col = "blue", 
     main = "SGD plot of alpha", xlab = "Iterations", ylab = "Estimate alpha")
abline(h = true_alpha, col = "red", lwd = 2, lty = 2)

# plot beta
plot(x = 0:5000, y = result_2$Path[,2], type = "l", col = "blue", 
     main = "SGD plot of beta", xlab = "Iterations", ylab = "Estimate beta") 
abline(h = true_beta, col = "red", lwd = 2, lty = 2)
```

**(d)**
We can observe that with a smaller sample size, there are larger fluctuations before converging. But generally, no matter the sample size and the starting point, the algorithm can still approximate the true parameters. We conclude that SGD’s ability to
converge despite using stochastic approximations to the gradient is good.


\newpage 

## QUESTION 4: Polynomial Regression with IRLS [23 points]

**(a)**
```{r}
lad_rho_prime <- function(residuals) {
  return(sign(residuals))
}
```

**(b)**
```{r}
#' Compute Residuals for a Polynomial Model.
#'
#' @param x A numeric vector of predictor values.
#' @param y A numeric vector of response values of the same length as `x`.
#' @param theta A numeric vector of coefficients for the polynomial model.
#'
#' @return A numeric vector of residuals, computed as the difference between
#'   the observed values `y` and the fitted values based on the polynomial model.
#'
#' @details Let
#'   ```
#'   poly_p(x, theta) = theta[1] + theta[2] * x + theta[3] * x^2
#'                      + ... + theta[p+1] * x^p
#'   ```
#'   denote a p-degree polynomial in `x` specified by the coefficients in `theta`.
#'   The function calculates the residuals for a model of this form, given the
#'   observed response values `y` and the fitted values
#'   ```
#'   fitted_values = poly_p(x, theta).
#'   ```
#'   The residuals are then computed as
#'   ```
#'   residuals = y - fitted_values.
#'   ```
getResids <- function(x, y, theta) {
  p <- length(theta) - 1
  fitted_values <- sapply(x, function(xi) sum(theta * xi^(0:p)))
  residuals <- y - fitted_values
  return(residuals)
}
```    

**(c)**
```{r}
library(ggplot2)
temperature_data <- read.csv("kw_temperature.csv")
temperature_data <- na.omit(temperature_data)
temperature_data$Day_Of_Year_Scaled <- temperature_data$Day_of_Year / 365

# Scatterplot of daily temperature data
ggplot(data = temperature_data, aes(x = Day_Of_Year_Scaled, y = Mean_Temp)) +
  geom_point(color = "blue", alpha = 0.5) +
  labs(title = "Daily Average Temperature in Kitchener-Waterloo",
       x = "Proportion of Year Elapsed",
       y = "Mean Temperature (°C)") +
  theme(plot.title = element_text(hjust = 0.5))

```

**(d)**
```{r}
getWeights <- function(resids, rhoPrimeFn, delta = 1e-12) {
    ## for calculating weights, minimum |residual| will be delta
    smallResids <- abs(resids) <= delta
    ## take care to preserve sign (in case rhoPrime not symmetric)
    resids[smallResids] <- delta * sign(resids[smallResids])
    ## calculate and return weights
    rhoPrimeFn(resids)/resids
}

getTheta <- function(y, x, wt, p) {
  # construct matrix X = [1, x, x^2, ..., x^p]
  N <- length(x)
  X <- matrix(1, nrow = N, ncol = p+1)
  for(i in 1:p) {
    X[, i+1] <- X[,i] * x
  }
  # compute WLS solution
  fit <- stats::lm.wfit(x = X, y = y, w = wt)
  # return fitted coefficients with names removed
  setNames(fit$coefficients, NULL) 
}

# Generalized function to compute residuals for a polynomial of degree p
getResids <- function(y, x, wt, theta) {
  p <- length(theta) - 1
  # Calculate fitted values based on polynomial terms
  fitted_values <- sapply(x, function(xi) sum(theta * xi^(0:p)))
  # Calculate residuals
  residuals <- y - fitted_values
  return(residuals)
}

# Function to format and print polynomial equations from coefficients
format_polynomial <- function(coefficients) {
  terms <- sapply(0:(length(coefficients) - 1), function(i) {
    coef <- coefficients[i + 1]
    # Format terms based on the degree
    if (i == 0) {
      sprintf("%.4f", coef)
    } else if (i == 1) {
      sprintf("%.3f*x", coef)
    } else {
      sprintf("%.3f*x^%d", coef, i)
    }
  })
  # Combine terms into a single equation string
  equation <- paste(terms, collapse = " + ")
  return(equation)
}

irls <- function(y, x, theta, rhoPrimeFn, 
                 dim = 2, delta = 1E-10,
                 testConvergenceFn = testConvergence,
                 maxIterations = 100,   # maximum number of iterations 
                 tolerance = 1E-6,      # parameters for the test
                 relative = FALSE       # for convergence function 
){if (missing(theta)) {theta <- rep(0, dim)}
  ## Initialize
  converged <- FALSE
  i <- 0
  N <- length(y)
  wt <- rep(1,N)
  ## LOOP
  while (!converged & i <= maxIterations) {
    ## get residuals
    resids <- getResids(y, x, wt, theta)
    ## update weights  (should check for zero resids)
    wt <- getWeights(resids, rhoPrimeFn, delta)
    ## solve the least squares problem
    thetaNew <- getTheta(y, x, wt,p)
    ##
    ## Check convergence
    converged <- testConvergenceFn(thetaNew, theta,
                                   tolerance = tolerance,
                                   relative = relative)
    ## Update iteration
    theta <- thetaNew
    i <- i + 1
  }
  ## Return last value and whether converged or not
  list(theta = theta, converged = converged, iteration = i)
}

# p_values for polynomial fitting
p_values <- 2:5

# List to store models
models <- list()

# Perform IRLS for each degree
for (p in p_values) {
  # Initialize theta with ones
  theta_initial <- rep(1, p + 1)
  
  # Run IRLS with custom getResids and lad_rho_prime
  model <- irls(
    y = temperature_data$Mean_Temp,
    x = temperature_data$Day_Of_Year_Scaled,
    theta = theta_initial,
    rhoPrimeFn = lad_rho_prime,
    dim = p + 1,
    delta = 1e-9,
    tolerance = 1e-6
  )
  
  models[[as.character(p)]] <- model$theta
  
  equation <- format_polynomial(model$theta)
  cat(sprintf("p=%d: y = %s\n", p, equation))
}
```
$$ $$

**(e)**
```{r, fig.width = 8}
library(ggplot2)

# Polynomial equations
temperature_data$y_p2 <- -16.6068 + 125.1373 * temperature_data$Day_Of_Year_Scaled + 
                          -113.4570 * temperature_data$Day_Of_Year_Scaled^2

temperature_data$y_p3 <- -7.8893 + 40.5960 * temperature_data$Day_Of_Year_Scaled + 
                          83.4903 * temperature_data$Day_Of_Year_Scaled^2 + 
                          -127.5908 * temperature_data$Day_Of_Year_Scaled^3

temperature_data$y_p4 <- -0.2268 + -112.5426 * temperature_data$Day_Of_Year_Scaled + 
                          777.3231 * temperature_data$Day_Of_Year_Scaled^2 + 
                          -1225.4857 * temperature_data$Day_Of_Year_Scaled^3 + 
                          559.6215 * temperature_data$Day_Of_Year_Scaled^4

temperature_data$y_p5 <- -1.3196 + -76.5246 * temperature_data$Day_Of_Year_Scaled + 
                          519.3577 * temperature_data$Day_Of_Year_Scaled^2 + 
                          -532.7390 * temperature_data$Day_Of_Year_Scaled^3 + 
                          -221.0394 * temperature_data$Day_Of_Year_Scaled^4 + 
                          311.9425 * temperature_data$Day_Of_Year_Scaled^5

# Plot with polynomial overlay
ggplot(data = temperature_data, aes(x = Day_Of_Year_Scaled, y = Mean_Temp)) +
  geom_point(color = "darkgrey", alpha = 0.6) +
  labs(title = "Polynomial Fits to Daily Average Temperature",
       x = "Proportion of Year Elapsed",
       y = "Mean Temperature (°C)",
       color = "Polynomial Degree") +
  # Add polynomial curves using calculated values
  geom_line(aes(y = y_p2, color = "2"), size = 0.9) +
  geom_line(aes(y = y_p3, color = "3"), size = 0.9) +
  geom_line(aes(y = y_p4, color = "4"), size = 0.9) +
  geom_line(aes(y = y_p5, color = "5"), size = 0.9) +
  theme_minimal() +
  labs(colour = " ") +
  theme(plot.title = element_text(hjust = 0.5))
```

**(f)**
```{r}
# Sample sizes for variability analysis
sample_sizes <- c(50, 100, 500)

# Set up the plotting layout
par(mfrow = c(1, 3), oma = c(0, 0, 2, 0))
par(mar = c(4, 4, 3, 1))

# Loop through each sample size
for (i in seq_along(sample_sizes)) {
  n <- sample_sizes[i]
  
  # Store fitted models for each sample
  sample_models <- list()
  
  # Generate 100 different samples for this sample size
  for (j in 1:100) {
    set.seed(i * 1000 + j)  # Unique seed for each iteration and sample size
    
    # Draw a random sample of size n
    sample_indices <- sample(seq_len(nrow(temperature_data)), size = n)
    sample_data <- temperature_data[sample_indices, ]
    
    # Fit the model
    model <- irls(
      y = sample_data$Mean_Temp,
      x = sample_data$Day_Of_Year_Scaled,
      theta = rep(1, 6),
      rhoPrimeFn = lad_rho_prime,
      dim = 6,
      delta = 1e-9,
      tolerance = 1e-6
    )
    sample_models[[j]] <- model$theta
  }
  
  # Create the plot using base R
  plot(temperature_data$Day_Of_Year_Scaled, temperature_data$Mean_Temp,
       pch = 20, col = adjustcolor("blue", alpha.f = 0.2), cex = 0.5,
       xlab = paste("Year Elapsed%, Sample Size =", n),
       ylab = "Mean Temperature (°C)")
  
  # Add sample fits
  x_seq <- seq(min(temperature_data$Day_Of_Year_Scaled), 
               max(temperature_data$Day_Of_Year_Scaled), 
               length.out = 100)
  
  # Add individual sample fits
  for (theta in sample_models) {
    y_pred <- sapply(x_seq, function(x) sum(theta * x^(0:5)))
    lines(x_seq, y_pred, col = adjustcolor("black", alpha.f = 0.3))
  }
  
  # Add the population quintic fit
  if (exists("models")) {
    y_pred_full <- sapply(x_seq, function(x) sum(models[["5"]] * x^(0:5)))
    lines(x_seq, y_pred_full, col = "red", lwd = 2)
  }
}

# Add an overall title
mtext("Sampling Variability in Quintic Polynomial Fits", 
      side = 3, outer = TRUE, line = -2)

# Reset the plotting parameters to default
par(mfrow = c(1, 1))
par(mar = c(5.1, 4.1, 4.1, 2.1))
```

**(g)**
Observe that when n = 50, variability is high in fitted curves, reflecting more sample error. When n =100, variability is reduced with better alignment to the population fit. When n = 500, variability is really small and closely approximating the population curve.
We can conclude that as sample sizes increases, sample error decreases, sampling variability decreases, and the regression estimation is more accurate.

